Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-27 doc/invt.tex
invtV7.1: small changes

The schedule, $\alpha(t)$,
also defines a variable transformation from
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited to the interval $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$,
as shown in Fig. \ref{fig:massq}.
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution decays exponentially,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,\footnote{If
  the magnitudes of the eigenvalues, $\lambda_k$'s,
  span a wide range, then for a large $\bar q$,
  the sum in Eq. \eqref{eq:mass_func} will be dominated by
  the largest term, where $\lambda_k \approx 1/\bar q$,
  and $M(\bar q) \approx \sqrt{\Gamma_k}/\bar q$.
  }
  %
then we would have an exponentially decaying schedule.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated
at a constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is typically true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.
}

@@

Many commonly-used updating schemes,
such as the Gaussian updating scheme
used in metadynamics,
can be characterized by a rigid window
centered on the current bin.
%
The eigenvalues of these translationally invariant or homogeneous
schemes can be readily deduced
from the window function ($\mu_0, \mu_1, \dots, \mu_b$)
as a cosine transform,
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ i = 1 }^b
  \mu_i
  \cos\left(
  \frac{ k \, i \, 2 \, \pi }
       {      g \, n        }
  \right)
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where $g = 1$ here for the periodic variable,
or $2$ for a non-periodic one
(cf. Appendix \ref{sec:more_wband}).
%
For the Gaussian updating scheme,
we have\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left[
        -\frac 1 2
        \left(
         \frac{ 2 \, \pi \, k \, \sigma_z }
              { g \, \Delta z  }
        \right)^2
      \right]
  ,
%\notag
\label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ and $\Delta z = z_{\max} - z_{\min}$
are the width of the Gaussian and the span
in the original collective variable, $z$,
respectively.

@@

\item \label{step:Gamma}
Estimate $\Gamma_k$'s from Eq. \eqref{eq:varXt}
(cf. Appendix \ref{sec:Gamma_measure})
as well as the initial bias $\epsilon_k$.\footnote{The
  value of $\tilde u_k$ can be estimated
  from the histogram-corrected bias potential
  by Eq. \eqref{eq:uhatav},
  although it leads to overestimated
  $\epsilon_k$ for large-$k$ modes.
}

\item \label{step:qT}
Compute the optimal value of $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
%
We implemented a variant of the Newton-Raphson method\cite{press3rd}
with the bisection method\cite{press3rd}
as a fallback.\footnote{
The equation was solved for $\ln q(T)$
to avoid negative values of $q(T)$,
and we imposed a lower bound,
$\bigl|\mathcal Y\bigl(q(T)\bigr)/q(T)\bigr|$, for
$\mathcal Y'\bigl(q(T)\bigr)$
in computing its change.
The initial value is set to $q(T) = \ln(1+T\,a_0/2)$.
}
%
%Since $m(\bar q)$ is a decreasing function of $\bar q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
  numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
  we have chosen the $\mathcal N+1$ grid points
  ($\mathcal N=20000$ here)
  over $[0, q(T)]$ to be given by
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
  where $i = 0, \dots, \mathcal N$.\footnote{For
    a mass distribution, $m(\bar q)$, that is
    inversely proportional to $\bar q$,
    this grid would make the integral,
    or the difference, $t_{[i+1]} - t_{[i]}$,
    roughly a constant.}
%
The curve $q(t)$ is then integrated
  using the trapezoidal rule\cite{press3rd},
  resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude by numerical differentiation:
  $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$\footnote{This
    discretization follows from runtime average interpretation
    discussed in Appendix \ref{sec:equilerr}.}
  and
  $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
  with
  $q_{[ -1]} \equiv 0$,
  $q_{[N+1]} \equiv q(T)$,
  $t_{[ -1]} \equiv 0$, and
  $t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
  can be linearly interpolated
  to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

The above procedure is not always necessary.
%
As shown below, for a class of bandpass updating schemes,
  which resemble the Gaussian updating schemes,
  the asymptotically optimal schedule
  is simply given by the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
%
%For these updating schemes,
%  the preliminary run with a constant updating magnitude
%  can be replaced by the WL-style stagewise reduction
%  until the updating magnitude falls under
%  the value given the inverse-time prescription\cite{
%    belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

Since an updating matrix
with identical eigenvalues
is a multiple of the identity matrix,\footnote{Since
  the eigenmode of $k = 0$ represents
  a uniform shift of the bias potential,
  we can freely set $\lambda_0$ to be
  the same as $\lambda_1$
  without changing the nature of
  the updating scheme.}
we recover the single-bin scheme
by setting the eigenvalues to $1$.
%
Thus, the single-bin scheme is most efficient
in error reduction in the asymptotic limit.
%
In particular,
with $\vartheta_k = (t+t_0)/(T+t_0)$,
we can readily recover the optimal schedule,
Eq. \eqref{eq:alpha_invt1}, with
the asymptotic error being
$\Err_A(T) = \sum_{ k = 1 }^{n-1} \Gamma_k \, T / (T + t_0)^2$,
%
%\begin{align}
%  \Err_A(T)
%  =
%  \frac{       T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{n-1}
%    \Gamma_k
%  ,
%\end{align}
%
which recovers Eq. \eqref{eq:EA_sbin1}.\footnote{By
Eq. \eqref{eq:parseval}, we have
$\sum_{k=1}^{n-1} \bigl\langle {\tilde \zeta}_k(t) \, {\tilde \zeta}_k(0) \bigr\rangle
= \sum_{i=1}^n \rho_i \bigl\langle \zeta_{*i}(t) \, \zeta_{*i}(0) \bigr\rangle$,
and by summing $t$ from $-\infty$ to $\infty$, we get
$\sum_{k=1}^{n-1} \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma$.}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that completely eliminates certain short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
accompanied by a perfect filter
for the designated noise modes,
which are assigned with zero eigenvalues.
%
For example,
by assuming
that the PMF is so smooth
that it can be represented by the first $K$
modes,\footnote{As an estimate of error,
  we note that truncating a linear or quadratic function
  at mode $K$ (for a non-periodic variable)
  results in an error roughly proportional to
  $|\max_i u_i - \min_i u_i|^2/K^3$.}
we demand
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0.
  \label{eq:lambda_bandpass}
\end{equation}
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%%
The asymptotic error is then
$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
which is less than that from Eq. \eqref{eq:EA_sbin1}.
%
The optimal schedule of these schemes
remains the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1}
with $t_0$ given by Eq. \eqref{eq:t0_sbin}.
%
%The asymptotic error is given by
%%
%\begin{equation}
%  \Err_A(T)
%  =
%  \frac{   T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{ K - 1 }
%    \Gamma_k
%  ,
%  \notag
%  %\label{eq:error_sinc}
%\end{equation}
%%
%as a generalization of Eq. \eqref{eq:Emin_sbin}
%for the single-bin updating scheme.
%

Since bandpass updating schemes
straightforward generalizations of
the single-bin updating scheme,
we can also borrow the WL algorithm
to control the updating magnitude
in initial stages of the simulation.
%
However, the histogram fluctuation should
be limited to first $K$ modes,
as the bandpass updating scheme
only partially flattens these histogram
modes (cf. Appendix \ref{sec:hfluc}).

@@

The PMF was calculated from $r = 0$ to half of the box size
(also the cutoff distance of the potential)
on a grid of size $\delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho^* = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
The reduced temperature was $T^* = 3$.
%
We tested the single-bin, Gaussian, and bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under a constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The equivalent $\ln f = n \, a_0$ for the WL algorithm
was $0.05$ and $0.025$, respectively.}
followed by a production phase of another $10^7$ steps
with updating magnitude given by
either the inverse-time or optimal schedule.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed
$\Gamma_k$'s and $\epsilon_k$'s,
although ideally they should be estimated from the data
collected from the equilibration phase in individual runs.

As shown in Table \ref{tab:lj_error},
all updating schemes
with either the inverse-time or optimal schedule
were able to reduced
the error significantly,
and the errors roughly matched the predicted values
calculated using the pre-computed
$\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian updating scheme produced less error than
the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
The effectiveness of the Gaussian updating scheme
in the low-density case
was probably because the sampling along
the collective variable was free from major barriers
and thus the histogram fluctuation largely
came from local noise.
%
%The optimal schedule also appeared to be more effective
%in reducing the error than the inverse-time schedule.
%%
In this case, we could further reduced the error
by using the bandpass updating scheme
(with the mode cutoff $K = 20$).
%
On the other hand, in the high-density case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range fluctuation,
$\Gamma_1 \approx \Gamma$,
and the reduction of local noise
had little effect on the total error.

@@

To illustrate the mode dependence of the error,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index, $k$,
for the $\varrho^* = 0.1$ case
with width $\sigma = 0.2$.
%
For modes $k=10$ to $25$,
the final errors from the Gaussian updating scheme
were greater than those from the single-bin updating scheme
although the initial values were less.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to
reduce the updating magnitude too quickly
leaving a hump on the error profile
for these middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
As a result, the optimal schedule usually produced
less error than the inverse-time schedule
as shown in Table \ref{tab:lj_error}.
%
Although the gain of using the optimal schedule
was modest in the reported cases,
it would increase with the simulation length.\footnote{
  For example, for the $\varrho^* = 0.1$ case
  with width $\sigma = 0.2$,
  the predicted error of a simulation of $10^8$ steps
  was $2.5\times10^{-5}$ for the inverse-time schedule
  or $3.7\times10^{-6}$ for the optimal schedule.}

>>>>

The schedule, $\alpha(t)$,
also defines a variable transformation from
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited to the interval $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$,
as shown in Fig. \ref{fig:massq}(b).
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution decays exponentially,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,\footnote{If
  the magnitudes of the eigenvalues, $\lambda_k$'s,
  span a wide range, then for a large $\bar q$,
  the sum in Eq. \eqref{eq:mass_func} will be dominated by
  the largest term, where $\lambda_k \approx 1/\bar q$,
  and $M(\bar q) \approx \sqrt{\Gamma_k}/\bar q$.
  }
  %
then we would have an exponentially decaying schedule.

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated
at a constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is often true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.
}

@@

Many commonly-used updating schemes,
such as the Gaussian updating scheme
used in metadynamics,
can be characterized by a rigid window
centered on the current bin.
%
If we denote the relative updating magnitude on
the $i$th bin from the center by $\mu_i$,
the eigenvalues of these translationally invariant or homogeneous
schemes can be readily deduced
from the window function ($\mu_0, \mu_1, \dots, \mu_b$)
as a cosine transform,
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ i = 1 }^b
  \mu_i
  \cos\left(
  \frac{ k \, i \, 2 \, \pi }
       {      g \, n        }
  \right)
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where $g = 1$ here for the periodic variable,
or $2$ for a non-periodic one
(cf. Appendix \ref{sec:more_wband}).
%
For the Gaussian updating scheme,
we have\cite{bussi2006}
\begin{align}
  \lambda_k
  =
  \exp\left[
        -\frac 1 2
        \left(
         \frac{ 2 \, \pi \, k \, \sigma_z }
              { g \, \Delta z  }
        \right)^2
      \right]
  ,
%\notag
\label{eq:lambda_Gaussian}
\end{align}
where $\sigma_z$ and $\Delta z = z_{\max} - z_{\min}$
are the width of the Gaussian and the span
in the original collective variable, $z$,
respectively.

@@

\item \label{step:Gamma}
Estimate $\Gamma_k$'s from Eq. \eqref{eq:varXt}
(cf. Appendix \ref{sec:Gamma_measure})
as well as the initial bias $\epsilon_k$.\footnote{The
  value of $\tilde u_k$ can be estimated
  from the histogram-corrected bias potential
  by Eq. \eqref{eq:uhatav},
  although it leads to overestimated
  $\epsilon_k$'s for large-$k$ modes.
}

\item \label{step:qT}
Compute the optimal value of $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
%
We implemented a variant of the Newton-Raphson method\cite{press3rd}
with the bisection method\cite{press3rd}
as a fallback.\footnote{
The equation was solved for $\ln q(T)$
to avoid negative values of $q(T)$,
and we imposed a lower bound,
$\bigl|\mathcal Y\bigl(q(T)\bigr)/q(T)\bigr|$, for
$\mathcal Y'\bigl(q(T)\bigr)$
in computing its change.
The initial value was set to $q(T) = \ln(1+T\,a_0/2)$.
}
%
%Since $m(\bar q)$ is a decreasing function of $\bar q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
  numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
  we have chosen the $\mathcal N+1$ grid points
  ($\mathcal N=20000$ here)
  over $[0, q(T)]$ to be given by
  $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
  where $i = 0, \dots, \mathcal N$.\footnote{For
    a mass distribution, $m(\bar q)$, that is
    inversely proportional to $\bar q$,
    this grid would make the integral,
    or the difference, $t_{[i+1]} - t_{[i]}$,
    roughly a constant.}
%
The curve $q(t)$ is then integrated
  using the trapezoidal rule\cite{press3rd},
  resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude by numerical differentiation:
  $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$\footnote{This
    discretization follows from the runtime average interpretation
    discussed in Appendix \ref{sec:equilerr}.}
  and
  $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
  with
  $q_{[ -1]} \equiv 0$,
  $q_{[N+1]} \equiv q(T)$,
  $t_{[ -1]} \equiv 0$, and
  $t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
  can be linearly interpolated
  to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

The above procedure is not always necessary.
%
As we will show below, for a class of bandpass updating schemes,
  the asymptotically optimal schedule
  is simply given by the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
%
%For these updating schemes,
%  the preliminary run with a constant updating magnitude
%  can be replaced by the WL-style stagewise reduction
%  until the updating magnitude falls under
%  the value given the inverse-time prescription\cite{
%    belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

@@

Since an updating matrix
with identical eigenvalues
is a multiple of the identity matrix,\footnote{Since
  the $k = 0$ eigenmode represents
  a uniform shift of the bias potential,
  we can freely set $\lambda_0$ to be
  the same as $\lambda_1$
  without changing the nature of
  the updating scheme.}
we recover the single-bin scheme
by setting the eigenvalues to $1$.
%
Thus, the single-bin scheme is most efficient
in error reduction in the asymptotic limit.
%
In particular,
with $\vartheta_k = (t+t_0)/(T+t_0)$,
we can readily recover the optimal schedule,
Eq. \eqref{eq:alpha_invt1}, with
the asymptotic error being
$\Err_A(T) = \sum_{ k = 1 }^{n-1} \Gamma_k \, T / (T + t_0)^2$,
%
%\begin{align}
%  \Err_A(T)
%  =
%  \frac{       T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{n-1}
%    \Gamma_k
%  ,
%\end{align}
%
which recovers Eq. \eqref{eq:EA_sbin1}.\footnote{By
Eq. \eqref{eq:parseval}, we have
$\sum_{k=1}^{n-1} \bigl\langle {\tilde \zeta}_k(t) \, {\tilde \zeta}_k(0) \bigr\rangle
= \sum_{i=1}^n \rho_i \bigl\langle \zeta_{*i}(t) \, \zeta_{*i}(0) \bigr\rangle$,
and by summing $t$ from $-\infty$ to $\infty$, we get
$\sum_{k=1}^{n-1} \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma$.}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that completely eliminates certain short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
accompanied by a perfect filter
for the designated noise modes,
which are assigned with zero eigenvalues.
%
For example,
by assuming
that the PMF is so smooth
that it can be represented by the first $K$
modes,\footnote{As an estimate of error,
  we note that truncating a linear or quadratic function
  at mode $K$ for a non-periodic variable
  results in an error roughly proportional to
  $|\max_i u_i - \min_i u_i|^2/K^3$.}
we demand
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0.
  \label{eq:lambda_bandpass}
\end{equation}
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%%
The asymptotic error is then
$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
which is less than that from Eq. \eqref{eq:EA_sbin1}.
%
The optimal schedule of these schemes
remains the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1}
with $t_0$ given by Eq. \eqref{eq:t0_sbin}.
%
%The asymptotic error is given by
%%
%\begin{equation}
%  \Err_A(T)
%  =
%  \frac{   T     }
%       { (T + t_0)^2 }
%  \sum_{ k = 1 }^{ K - 1 }
%    \Gamma_k
%  ,
%  \notag
%  %\label{eq:error_sinc}
%\end{equation}
%%
%as a generalization of Eq. \eqref{eq:Emin_sbin}
%for the single-bin updating scheme.
%

Since bandpass updating schemes are
straightforward generalizations of
the single-bin updating scheme,
we can also borrow the WL algorithm
to control the updating magnitude
in initial stages of the simulation.
%
However, the histogram fluctuation should
be limited to first $K$ modes,
as the bandpass updating scheme
only partially flattens these histogram
modes (cf. Appendix \ref{sec:hfluc}).

@@

The PMF was calculated from $r = 0$ to half of the box size
(also the cutoff distance of the potential)
on a grid of size $\delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho^* = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
The reduced temperature was $T^* = 3$.
%
We tested the single-bin, Gaussian, and bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under a constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The equivalent $\ln f = n \, a_0$ for the WL algorithm
were $0.05$ and $0.025$ for $\varrho^* = 0.1$ and $\varrho^* = 0.8$, respectively.}
followed by a production phase of another $10^7$ steps
with updating magnitude given by
either the inverse-time or optimal schedule.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed
$\Gamma_k$'s and $\epsilon_k$'s,
although ideally they could be estimated from the data
collected from the equilibration phase in individual runs.

As shown in Table \ref{tab:lj_error},
all updating schemes
with either the inverse-time or optimal schedule
were able to reduced
the error significantly,
and the errors roughly matched the predicted values
calculated using the pre-computed
$\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian updating scheme produced less error than
the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
The effectiveness of the Gaussian updating scheme
in the low-density case
was probably because the sampling along
the collective variable was free from major barriers
and thus the histogram fluctuation largely
came from local noise.
%
%The optimal schedule also appeared to be more effective
%in reducing the error than the inverse-time schedule.
%%
In this case, we could further reduced the error
by using the bandpass updating scheme
(with the mode cutoff $K = 20$).
%
On the other hand, in the high-density case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range fluctuation,
as $\Gamma_1 \approx \Gamma$,
and the reduction of local noise
had little effect on the total error.

@@

To illustrate the mode dependence of the error,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index, $k$,
for the $\varrho^* = 0.1$ case
with width $\sigma = 0.2$.
%
For modes $k=10$ to $25$,
the final errors from the Gaussian updating scheme
were greater than those from the single-bin updating scheme,
although the initial values were less.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to
reduce the updating magnitude too quickly,
leaving a hump on the error profile
for these middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
As a result, the optimal schedule usually produced
less error than the inverse-time schedule,
as shown in Table \ref{tab:lj_error}.
%
Although the gain of using the optimal schedule
was modest in the reported cases,
it would increase with the simulation length.\footnote{
  For example, for the $\varrho^* = 0.1$ case
  with width $\sigma = 0.2$,
  the predicted error of a simulation of $10^8$ steps
  was $2.5\times10^{-5}$ for the inverse-time schedule
  or $3.7\times10^{-6}$ for the optimal schedule.}

One way to overcome the systematic bias
in the Gaussian updating scheme
is to use the histogram-based correction formula,
Eq. \eqref{eq:uhatav}.
%
As shown in Fig. \ref{fig:lj_xerr},
even for the simulation using the non-optimal $1/t$ schedule,
the correction was able to reproduce
the flatter profile seen in the single-bin case.
%
However, this correction also introduced
some local noise in the large $k$ modes,
and it would be best used as a post-simulation
analysis tool for simulations
with detectable residual bias.
%
For example, for the simulation in the $\varrho^* = 0.8$ case
with $\sigma = 0.5$ and the inverse-time schedule,
the correction was able to lower
the error from $9.8\times10^{-3}$
(Table \ref{tab:lj_error}) to $3.3\times10^{-3}$.
